---
output:
  html_document:
    toc: true
    toc_depth: 5
    toc_float: true
---
```{r intro, echo=FALSE, results="hide"}
knitr::opts_chunk$set(echo=FALSE, 
                      message=FALSE, 
                      comment = "", 
                      warning=FALSE, 
                      results="hide") 
knitr::opts_knit$set(root.dir = "C:/Users/YoonJoung Choi/Dropbox/0 Project/COVID19_US/")

date<-as.Date(Sys.time(	), format='%d%b%Y')
time<-Sys.time()

suppressWarnings(suppressMessages(library(dplyr)))
suppressWarnings(suppressMessages(library(ggplot2)))
suppressWarnings(suppressMessages(library(tidyverse)))
suppressWarnings(suppressMessages(library(plotly)))
suppressWarnings(suppressMessages(library(Matrix)))
suppressWarnings(suppressMessages(library(stringr)))
suppressWarnings(suppressMessages(library(stringi)))


suppressWarnings(suppressMessages(library(rlist)))
suppressWarnings(suppressMessages(library(zoo)))
suppressWarnings(suppressMessages(library(RColorBrewer))) 
suppressWarnings(suppressMessages(library(lubridate)))
```

###Updated results for a pre-print research note, _"Deaths among COVID Cases in the United States: Racial and Ethnic Disparities Persist"_

Madeleine Short Fabic, Yoonjoung Choi, David Bishai

(Updated: `r time` EDT)  

* The pre-print was published on November 17, 2020, and is available here: [https://www.medrxiv.org/content/10.1101/202011.15.20232066v2](https://www.medrxiv.org/content/10.1101/2020.11.15.20232066v2).   
* Data source: [COVID-19 Case Surveillance Public Use Data](https://data.cdc.gov/Case-Surveillance/COVID-19-Case-Surveillance-Public-Use-Data/vbim-akqf) (version: December 4, 2020, release)    

```{r dataPop}
# https://www.census.gov/newsroom/press-kits/2020/population-estimates-detailed.html
# Annual Estimates of the Resident Population by Sex, Age, Race, and Hispanic Origin for the United States: April 1, 2010 to July 1, 2019
# NC-EST2019-ASR6H

dtapop<-read.csv("C:/Users/YoonJoung Choi/Dropbox/0 Project/COVID19_US/DataCensusBureau/nc-est2019-asr6h-2019.csv")
names(dtapop)<- tolower(names(dtapop))
    colnames(dtapop) 
    table(dtapop$race) 
    table(dtapop$hispanicorigin) 
 
    #View(dtapop)

dtapopage<-dtapop%>%
    filter(sex=="Total")%>%
    filter(race=="Total")%>%
    filter(hispanicorigin=="Total")%>%
    select(race, hispanicorigin, sex, starts_with("age"), pop)%>%
    filter(agegroup_5=="yes")

    sum(dtapopage$pop)

dtapopage10<-dtapopage%>%
    mutate(
        age=ifelse(age== 5,  0, age),
        age=ifelse(age==15, 10, age),
        age=ifelse(age==25, 20, age),
        age=ifelse(age==35, 30, age),
        age=ifelse(age==45, 40, age),
        age=ifelse(age==55, 50, age),
        age=ifelse(age==65, 60, age),
        age=ifelse(age==75, 70, age),
        age=ifelse(age==85, 80, age)
    )%>%
    group_by(race, hispanicorigin, sex, age)%>%
    summarise_at(vars(pop), funs(sum))
    
    sum(dtapopage10$pop)

dtapopagerace<-dtapop%>%
    filter(sex=="Total")%>%
    filter(race!="Total")%>%
    filter(hispanicorigin!="Total")%>%
    select(race, hispanicorigin, sex, starts_with("age"), pop)%>%
    filter(agegroup_5=="yes")%>%
    mutate(
        racegroup="", 
        racegroup=ifelse((race=="White" & hispanicorigin=="Not Hispanic"),
                         "NH White", racegroup), 
        racegroup=ifelse((race=="Black" & hispanicorigin=="Not Hispanic"),
                         "NH Black", racegroup), 
        racegroup=ifelse((race=="Asian" & hispanicorigin=="Not Hispanic"),
                         "NH Asian", racegroup), 
        racegroup=ifelse((race=="AIAN" & hispanicorigin=="Not Hispanic"),
                         "NH AIAN", racegroup), 
        racegroup=ifelse((race=="NHPI" & hispanicorigin=="Not Hispanic"),
                         "NH NHPI", racegroup), 
        racegroup=ifelse((race=="TwoOrMoreRaces" & hispanicorigin=="Not Hispanic"),
                         "NH Multiple/Other", racegroup), 
        racegroup=ifelse((hispanicorigin=="Hispanic"),
                         "Hispanic", racegroup)
    )%>%
    select(racegroup, starts_with("age"), pop)%>%
    group_by(racegroup, age, agegroup)%>%
    summarise_at(vars(pop), funs(sum))%>%
    ungroup()

    sum(dtapopagerace$pop)
    
dtapopagerace10<-dtapopagerace%>%
    mutate(
        age=ifelse(age== 5,  0, age),
        age=ifelse(age==15, 10, age),
        age=ifelse(age==25, 20, age),
        age=ifelse(age==35, 30, age),
        age=ifelse(age==45, 40, age),
        age=ifelse(age==55, 50, age),
        age=ifelse(age==65, 60, age),
        age=ifelse(age==75, 70, age),
        age=ifelse(age==85, 80, age)
    )%>%
    group_by(racegroup, age)%>%
    summarise_at(vars(pop), funs(sum))    

    sum(dtapopagerace10$pop)

```

```{r dtasurveillance}
# COVID-19 Case Surveillance Public Use (case level)
# https://data.cdc.gov/Case-Surveillance/COVID-19-Case-Surveillance-Public-Use-Data/vbim-akqf

    ##### Export to CSV
    
dtasurveillance<-read.csv("C:/Users/YoonJoung Choi/Dropbox/0 Project/COVID19_US/DataCDC/COVID-19_Case_Surveillance_Public_Use_Data.csv")    
    names(dtasurveillance)<- tolower(names(dtasurveillance))
    colnames(dtasurveillance)
    nrow(dtasurveillance)
```

```{r dtacase}
# CASE LEVEL data managment         
dtacase<-dtasurveillance%>%
    mutate_if(is.factor, as.character) %>% 
    mutate(
        
        cdc_report_dt=as.Date(cdc_report_dt),
        week=as.numeric(
            strftime(as.POSIXct(cdc_report_dt), format = "%V")),
        month=as.numeric(
            substr(as.character(as.Date(cdc_report_dt, format = "%m")), 6, 7)), 
        
        sextotal="Total",
        
        agegroup=age_group, 
            agegroup=ifelse(is.na(agegroup)==TRUE, "Unknown", agegroup),
        age=as.numeric(sapply(strsplit(agegroup,"-"), `[`, 1)), #age interval begins
            age=ifelse(agegroup=="80+ Years", 80, age),
        
        racegroup=race.and.ethnicity..combined., 
        racegroup=ifelse(race.and.ethnicity..combined.=="White, Non-Hispanic",
                         "NH White", racegroup), 
        racegroup=ifelse(race.and.ethnicity..combined.=="Black, Non-Hispanic",
                         "NH Black", racegroup), 
        racegroup=ifelse(race.and.ethnicity..combined.=="Asian, Non-Hispanic",
                         "NH Asian", racegroup), 
        racegroup=ifelse(race.and.ethnicity..combined.=="American Indian/Alaska Native, Non-Hispanic",
                         "NH AIAN", racegroup), 
        racegroup=ifelse(race.and.ethnicity..combined.=="Native Hawaiian/Other Pacific Islander, Non-Hispanic",
                         "NH NHPI", racegroup), 
        racegroup=ifelse(race.and.ethnicity..combined.=="Multiple/Other, Non-Hispanic",
                         "NH Multiple/Other", racegroup), 
        racegroup=ifelse(race.and.ethnicity..combined.=="Hispanic/Latino",
                         "Hispanic", racegroup),
        racegroup=ifelse(is.na(race.and.ethnicity..combined.)==TRUE,
                         "Unknown", racegroup),
        
        raceagegroup=paste0(racegroup, "_",agegroup), 

        missingage  =as.numeric(is.na(age)==TRUE),
        missingsex  =as.numeric(sex!="Male" & sex!="Female" ),
        missingrace =as.numeric(racegroup=="Unknown"),
        missingdeath=as.numeric(death_yn!="Yes" & death_yn!="No"),  
        missinghosp =as.numeric(hosp_yn!="Yes" & hosp_yn!="No"),
        
        case=1, 
        death=as.numeric(death_yn=="Yes"), 
        hosp =as.numeric(hosp_yn=="Yes"), 
        
            death=ifelse(missingdeath==1, NA, death), 
            hosp =ifelse(missinghosp ==1, NA, hosp)
        
        )

colnames(dtacase)    
str(dtacase)
#summary(dtacase)

#table(dtacase$raceagegroup)
```

```{r dtamonthly}
# Weekly trend count data by RACE AND AGE
dtamonthly<-dtacase%>%
    arrange(racegroup, agegroup, month)%>%
    mutate(raceagegroupmonth=paste0(racegroup,"_",agegroup,"_",as.character(month)))%>%
    group_by(raceagegroupmonth)%>%
    summarise_at(vars(case, death, hosp, missingdeath, missinghosp), funs(sum), na.rm = TRUE)%>%
    ungroup()%>%
    mutate(
        racegroup=sapply(strsplit(raceagegroupmonth,"_"), `[`, 1),
        agegroup=sapply(strsplit(raceagegroupmonth,"_"), `[`, 2),
        month=as.numeric(sapply(strsplit(raceagegroupmonth,"_"), `[`, 3)), 
        
        age=as.numeric(sapply(strsplit(agegroup,"-"), `[`, 1)), #age interval begins
            age=ifelse(agegroup=="80+ Years", 80, age), 
        
        raceagegroup=paste0(racegroup, "_", agegroup) 
        )%>%
    arrange(racegroup, agegroup, month)
    

    #View(dtamonthly)
    #nrow(dtamonthly)
    #str(dtamonthly)
    #summary(dtamonthly)

dim(dtamonthly)
dtamonthly<-left_join(dtamonthly, dtapopagerace10, by = c("racegroup", "age")) 
dim(dtamonthly)    
```

```{r dtamonthly_standardization}
dta<-dtamonthly%>%
    mutate(
        releasedate="December 4, 2020", #"date of CDC public data release" , 
        totalcases=sum(case), 
        totaldeaths=sum(death), 
        arscfr = 100*death/case 
        #arscfr = ifelse(case<25, NA, arscfr) #"Age-Race-specific CFR by month of case report"
    )%>%
    group_by(agegroup)%>%
    mutate(standardcase=sum(case))%>%
    ungroup()%>%
    
#	********** Age-adjusted CFR 
#	*- ONLY March through september for November release  
#	*- ONLY March through october for December release      
#	*- ONLY among cases with both known race AND age => overall cases serves as standard    
    filter(month>=3 & month<=9)%>%
    filter(agegroup!="Unknown" & racegroup!="Unknown" )%>%
    mutate(
        death_adj = standardcase * arscfr /100 #/*number of deaths, based on standard population & ARS-CFR*/
    )%>%
    group_by(month, racegroup)%>%
    mutate(
    	case_monthly_observed  = sum(case),	
    	death_monthly_observed  = sum(death), 
    	
    	case_monthly_std  = sum(standardcase), 
    	death_monthly_std  = sum(death_adj) 
    )%>%
    ungroup()%>%
    mutate(
    	cfr_observed = 100* death_monthly_observed / case_monthly_observed,
    	cfr_std = 100* death_monthly_std / case_monthly_std        
    )

#dim(dta_annex)
```

```{r dtasummary}
dtamonthly_AARSCFR<-dta%>%
    select(month, racegroup, ends_with("_observed"), ends_with("_std") )%>%
    group_by(month, racegroup)%>%
    summarize_all(funs(mean(., na.rm=TRUE)))%>%
    ungroup()

#dim(dtamonthly_AARSCFR)
```

####__Figure 1. Trends of monthly case fatality rates among COVID-19 cases by race/ethnicity: crude and age-adjusted__     
NH: Non-Hispanic. Results for four largest single race/ethnicity groups are presented. Monthly case fatality rates refer to percent of cases reported in each month that have resulted in death either during the month or later.

```{r figure1, results="asis", fig.align="left", out.width="800px", out.height="600px"}

dtafig<-dtamonthly_AARSCFR%>%
    filter(
        racegroup=="Hispanic" | 
        racegroup=="NH Asian" |
        racegroup=="NH Black" |    
        racegroup=="NH White" )%>%
    mutate(
        racegroup=str_replace(racegroup, fixed(" "), "")
    )%>%
    mutate_at(vars(starts_with("cfr_")), 
              funs(round(., 1)) ) %>%
    select(month, racegroup, starts_with("cfr_") ) %>%
    gather(var, val, starts_with("cfr_") ) %>% 
    unite(var2, racegroup, var) %>% 
    spread(var2, val) %>%
    ungroup() #THIS LINE IS CRITICAL, BUT I DON"T KNOW WHY"

fig1<-dtafig%>%
    plot_ly(x = ~month, showlegend=FALSE)%>%
    add_trace( 
        y = ~Hispanic_cfr_observed, name="Hispanic",
        mode = 'markers+lines', 
        marker = list(color = 'blue'),
        line = list(color = 'blue') ) %>%
    add_trace(
        y = ~NHAsian_cfr_observed,  name="NH Asian",   
        mode = 'markers+lines', 
        marker = list(color = 'red'),
        line = list(color = 'red') ) %>%
    add_trace(
        y = ~NHBlack_cfr_observed,  name="NH Black",
        mode = 'markers+lines', 
        marker = list(color = "green"),
        line = list(color = "green") ) %>%
    add_trace(
        y = ~NHWhite_cfr_observed,  name="NH Black",
        mode = 'markers+lines', 
        marker = list(color = "black"),
        line = list(color = "black") ) %>%
    add_annotations(
        text = "Observed",
        x = 0.5, xref = "paper", xanchor = "center", 
        y = 0.9, yref = "paper", yanchor = "bottom", 
        showarrow = FALSE,
        font = list(size = 12) ) %>%    
    layout(
        showlegend = FALSE,
        title = c(""),
        xaxis=list(title="Month", range=c(2,10), showgrid = FALSE, 
                    tickfont = list(size=10)), 
        yaxis=list(title="Case-fatality rate (%)", showgrid = FALSE) 
        )

fig2<-dtafig%>%
    plot_ly(x = ~month)%>%
    add_trace( 
        y = ~Hispanic_cfr_std, name="Hispanic",
        mode = 'markers+lines', 
        marker = list(color = 'blue'),
        line = list(color = 'blue') ) %>%
    add_trace(
        y = ~NHAsian_cfr_std,  name="NH Asian",   
        mode = 'markers+lines', 
        marker = list(color = 'red'),
        line = list(color = 'red') ) %>%
    add_trace(
        y = ~NHBlack_cfr_std,  name="NH Black",
        mode = 'markers+lines', 
        marker = list(color = "green"),
        line = list(color = "green") ) %>%
    add_trace(
        y = ~NHWhite_cfr_std,  name="NH Black",
        mode = 'markers+lines', 
        marker = list(color = "black"),
        line = list(color = "black") ) %>%
    add_annotations(
        text = "Age-standardized",
        x = 0.5, xref = "paper", xanchor = "center", 
        y = 0.9, yref = "paper", yanchor = "bottom", 
        showarrow = FALSE,
        font = list(size = 12) ) %>%    
    layout(
        title = c(""),
        xaxis=list(title="Month", range=c(2,10), showgrid = FALSE, 
                    tickfont = list(size=10)), 
        yaxis=list(title="Case-fatality rate (%)", showgrid = FALSE) 
        )

subplot(fig1, fig2, margin=0.05, shareY = TRUE , titleX = TRUE)%>%
    layout(
        title = c(""),
        xaxis=list(title="Month", showgrid = FALSE), 
        legend=list(orientation="h", font=list(size=10), 
                xanchor = "center", x = 0.5, 
                yanchor = "center", y = -0.1)
        )

```

####__Figure 2. Monthly age-specific case fatality rates across race/ethnicity groups__    
NH: Non-Hispanic. Results for four largest single race/ethnicity groups are presented. Monthly case fatality rates refer to percent of cases reported in each month that have resulted in death either during the month or later.    

```{r figure2, results="asis", fig.align="left", out.width="800px", out.height="600px"}

dtafig<-dta%>%
    filter(age>=30)%>%
    filter(
        racegroup=="Hispanic" | 
        racegroup=="NH Asian" |
        racegroup=="NH Black" |    
        racegroup=="NH White" )%>%
    mutate(
        racegroup=str_replace(racegroup, fixed(" "), "")
    )%>%
    mutate_at(vars(arscfr), 
              funs(round(., 1)) ) %>%
    select(month, agegroup, racegroup, arscfr) %>%
    gather(var, val, arscfr ) %>% 
    unite(var2, racegroup, var) %>% 
    spread(var2, val) %>%
    ungroup() #THIS LINE IS CRITICAL, BUT I DON"T KNOW WHY"

panel <- . %>% 
    plot_ly(x = ~month)%>% 
    add_trace( 
        y = ~Hispanic_arscfr, name="Hispanic",
        mode = 'markers+lines', 
        marker = list(color = 'blue'),
        line = list(color = 'blue') ) %>%
    add_trace(
        y = ~NHAsian_arscfr,  name="NH Asian",   
        mode = 'markers+lines', 
        marker = list(color = 'red'),
        line = list(color = 'red') ) %>%
    add_trace(
        y = ~NHBlack_arscfr,  name="NH Black",
        mode = 'markers+lines', 
        marker = list(color = "green"),
        line = list(color = "green") ) %>%
    add_trace(
        y = ~NHWhite_arscfr,  name="NH Black",
        mode = 'markers+lines', 
        marker = list(color = "black"),
        line = list(color = "black") ) %>%
    add_annotations(
        text = ~unique(agegroup),
        x = 0.5, y = 0.90, xref = "paper", yref = "paper",    
        xanchor = "center", yanchor = "bottom", showarrow = FALSE,
        font = list(size = 12)  )%>%
    layout(
        showlegend=FALSE, 
        title = c(""),
        xaxis=list(title="Month", range=c(2,10), showgrid = FALSE, 
                    tickfont = list(size=10)), 
        yaxis=list(title="Case-fatality rate (%) (log scale)", 
                   type="log", 
                   showgrid = FALSE)
        )

dtafig%>%
    group_by(agegroup) %>%
    do(p = panel(.)) %>%
    subplot(nrows = 3, shareX = TRUE, shareY = FALSE)  

```

---

<p style="color:lightgray">
See [GitHub](https://github.com/yoonjoung/COVID19_US_AACFR) for reproducible code.    
For typos, errors, and questions, contact me at [www.isquared.global](https://www.iSquared.global/YJ).   

_Making Data Delicious, One Byte at a Time_, in good times and bad times.</p>